\thornado-hydro: Generalizing discontinuous galerkin methods for a nuclear equation of state for supernova hydrodynamics

### 2019-03-07 paper/ms.tex
Right and Left eigenvectors reduced and added to manuscript. Old notebooks deleted.

\documentclass[onecolumn]{aastex62}
\usepackage{times}
\usepackage{epsfig}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{color}
\usepackage{microtype}
\usepackage{float}
\usepackage{stfloats}
\usepackage{natbib}
\usepackage{verbatim}
\usepackage{placeins}
\graphicspath{{./figures/}}
\usepackage{wrapfig}
\hypersetup{backref,breaklinks,colorlinks,citecolor=blue}
\usepackage[all]{hypcap}
\usepackage{definitions}

@@

%\begin{abstract}
%\end{abstract}
\begin{abstract}
\todo{Fill In Abstract}
A problem of high importance in computational astrophysics is obtaining accurate
solutions to the Euler equations of hydrodynamics. We are interested in solving
the Euler equations in the context of core collapse supernovae (CCSNe).
The toolkit for high-order neutrino-radiation hydrodynamics (\thornado)
is being developed for CCSNe simulations and related problems utilizing a spatial discretization
based on the discontinuous Galerkin (DG) method.
Because the Euler equations form a hyperbolic set of partial differential equations,
they may be linearized to advection equations by the introduction of
chracteristic variables related to the eigen-decomposition.
This use of characteristic variables also increases the efficiency of the slope
limiting process. However, the addition of the nuclear equation of state equation
to the Euler equations makes the decompisition nontrivial.
We introduce the framework for the characteristic decomposition of the Euler equations
with the inclusion of the nuclear EOS terms and
present results from some initial tests.
\end{abstract}

@@

\section{Euler Equations of Gas Dynamics in Cartesian Coordinates}
\todo{Use this form of the Euler Equations?}
The non-relativistic Euler equations of gas dynamics
\citep[see, e.g.,][for details]{leveque:2002} in cartesian coordinates are given by
the equations of conservation of mass,
\beq
  \pderiv{\rho}{t} + \pderiv{}{x}(\rho u) = 0
  \label{eq:massConservation}
\eeq
conservation of momentum,
\beq
	\pderiv{(\rho u)}{t} + \pderiv{}{x}(\rho u^2 + p) = 0
  \label{eq:momentumConservation}
\eeq
conservation of energy,
\beq
  \pderiv{E}{t} + \pderiv{}{x}((E+p)u) = 0
  \label{eq:energyConservation}
\eeq
and conservation of \textcolor{red}{electron density}
\beq
  \pderiv{D}{t} + \pderiv{}{x}(D u) = 0
  \label{eq:electronConservation}
\eeq
where $\rho$ represents mass density, $u$ the fluid velocity, $p$ the fluid
pressure, $D=\rho y_e$ is the \textcolor{red}{what is this?},
and $E=e+\frac{1}{2}\rho u^2$ the total energy (internal plus kinetic).
The system of equations is closed by the EOS which links the pressure to
the quantities evolved by Equations
\eqref{eq:massConservation}-\eqref{eq:electronConservation}. The inclusion of
Equation \eqref{eq:electronConservation} is because we require a nuclear EOS.
We may rewrite Equations \eqref{eq:massConservation}-\eqref{eq:electronConservation}
in a more convenient way:

@@

\subsection{Slope Limiting}
A common feature of numerical ODEs is unphysical oscillations in the solutions.
It is therfore of great interest in the DG algorithm to implement slope limiting
of the polynomial $\mathbf{U}_h$. We use the total variation diminishing (TVD)
slope limiter \citep[see, e.g.,][]{cockburn:1998} in conjunction with the
troubled cell indicator discussed in \citet{fu:2017} to prevent excessive limiting.

The general $s$-stage Runge-Kutta time stepping algorithm, including
the limiting process, can then be summarized as in \citet{cockburn:2001}:
\begin{itemize}
  \item[1.] Set $\bar{\vect{U}}^{(0)} = \bar{\vect{U}}^{n}$,
  \item[2.] For $i=1,\ldots,s$ compute:
  \begin{equation}
  \small  \bar{\vect{U}}^{(i)}
    = \Lambda^{\Pos}\Big\{\Lambda^{\TVD}\Big\{\sum_{j=0}^{i-1}\alpha_{ij}\,\bar{\vect{U}}^{(j)}+\beta_{ij}\,\Delta t\,\bar{\vect{F}}\big(\bar{\vect{U}}^{(j)}\big)\Big\}\Big\},
    \label{eq:rkStages}
  \end{equation}
  \item[3.] Set $\bar{\vect{U}}^{n+1}=\bar{\vect{U}}^{(s)}$.
\end{itemize}
Above, the TVD limiter and the `positivity' limiter preventing unphysical
states are denoted by the operators $\Lambda^{\TVD}\{\}$
and $ \Lambda^{\Pos}\{\}$, respectively. The SSP-RK3 coefficients $\alpha_{ij}$
and $\beta_{ij}$ may be found in Table~2.1 in \citet{cockburn:2001}.

@@

\beq
  \pderiv{\mathbf{F(\mathbf{U})}}{\mathbf{U}} =
  \mathcal{R} \Lambda \mathcal{R}^{-1},
\eeq
where the columns of $R$ contain the right eigenvectors of the Jacobian,
the rows of $R^{-1}$ contain the left eigenvectors, and
$\Lambda$ is a diagonal matrix containing the eigenvalues of the Jacobian.
At this point, we will introduce the characteristic variable
$\mathbf{w} = R^{-1}\mathbf{U}$. Multiplying both sides of Equation
\eqref{eq:charEq} by $R^{-1}$, we are able to linearize the system of equations to
a system of advection equations

@@

\begin{table*}[ht!]

\begin{align}
  \centering
	\pderiv{\mathbf{F}(\mathbf{U})}{\mathbf{U}}
	= \left[
		\begin{array}{cccccc}
			0 & 1 & 0 & 0 & 0 & 0 \\
			-v_{1}^{2} -p_{\tau}\tau^{2} - p_{\epsilon}\tau(\epsilon - \frac{v^{2}}{2}) & v_{1}(2-p_{\epsilon}\tau)  & -p_{\epsilon}v_{2}\tau & -p_{\epsilon}v_{3}\tau  & p_{\epsilon}\tau  & p_{n} \\
			-v_{1}v_{2} & v_2 & v_1 & 0 & 0 & 0 \\
			-v_{1}v_{3} & v_{3} & 0 & v_{1} & 0 & 0 \\
			v_{1}(-H - p_{\tau}\tau^{2} -p_{\epsilon}\tau(\epsilon - \frac{v^{2}}{2})) & H - p_{\epsilon}v_{1}^{2}\tau  & -p_{\epsilon}v_{1}v_{2}\tau & -p_{\epsilon}v_{1}v_{3}\tau  & v_{1}(1+p_{\epsilon}\tau) & v_{1} p_{n} \\
			-v_{1} Y_{e}/m_{B} & Y_{e}/m_{B} & 0 & 0 & 0 & v_{1} \\
		\end{array}
    \right]
\end{align}
\caption{}
\end{table*}

@@

\begin{align}
\Lambda =
\begin{bmatrix}
  v_{1} - c_{s} & 0 & 0& 0& 0& 0\\
  0 & v_{1} & 0 & 0 & 0 & 0 \\
  0 & 0 & v_{1} & 0 & 0 & 0 \\
  0 & 0 & 0 & v_{1} & 0 & 0 \\
  0 & 0 & 0 & 0 & v_{1} & 0 \\
  0 & 0 & 0 & 0 & 0 & v_{1} + c_{s}
\end{bmatrix}
\end{align}
where $c_{S} = \sqrt{\tau (n p_{n} + p p_{\epsilon}\tau - p_{\tau}\tau)}$ is
the local sound speed. The right eigenvectors are then given by
\begin{align*}
  \mathcal{R}_{1} =
  \left[
  \begin{array}{cccccc}
   1 & 0 & 1 & 1 & 0 & 1 \\
   v_{1}-c & 0 & v_{1} & v_{1} & 0 & c+v_{1} \\
   v_{2} & 1 & 0 & 0 & 0 & v_{2} \\
   v_{3} & 0 & 0 & 0 & 1 & v_{3} \\
   h-c v_{1} & v_{2} & \beta & 0 & v_{3} & h+c v_{1} \\
   n \tau  & 0 & 0 & \frac{\tau  \chi }{2 p_{n}} & 0 & n \tau  \\
  \end{array}
  \right]
\end{align*}
where the follow definitions have been used:
$h_{n} = \frac{c^2}{p_{\epsilon}\tau} + k $, $k = \frac{-n p_{n}
+ p_{\epsilon} (\frac{1}{2}v^2 + \epsilon) + p_{\tau}\tau}{p_{\epsilon}}$,
$\delta_{1} = v_{1}^{2}-v_{2}^{2}-v_{3}^{2}$,
$\chi = p_{\epsilon} ( \delta_{1} + 2\epsilon) + 2p_{\tau}\tau$, and
$\beta = \frac{1}{2} (\delta_{1}+2 \epsilon +\frac{2 p_{\tau} \tau }{p_{\epsilon}})$.
\begin{align*}
  \mathcal{R}_{1}^{-1} = \frac{1}{c^2}
  \left[
  \begin{array}{cccccc}
   \frac{1}{4} (2 c  v_{1}+\omega ) & \frac{1}{2} (-c- \phi_{1} ) & -\frac{1}{2} \phi_{2}
    & -\frac{1}{2} \phi_{3}  & \frac{p_{\epsilon} \tau }{2} & \frac{p_{n}}{2}
     \\
   -\frac{v_{2} \omega }{2} & \phi_{1} v_{2}  & c^2+\phi_{2} v_{2}  &
     \phi_{2} v_{3}  & -\phi_{2}  & -p_{n} v_{2} \\
   \frac{2 \chi  c^2+\alpha  \omega }{2 \chi } & -\frac{\phi_{1} \alpha  }{\chi } &
     -\frac{\phi_{2} \alpha  }{\chi } & -\frac{\phi_{3} \alpha }{\chi } &
     \frac{p_{\epsilon} \alpha  \tau }{\chi } & \frac{p_{n} \left(\alpha  \tau -2 c^2\right)}{\tau  \chi } \\
   -\frac{n p_{n} \omega }{\chi } & \frac{2 n p_{n} \phi_{1} }{\chi } & \frac{2 n p_{n}
     \phi_{2} }{\chi } & \frac{2 n p_{n} \phi_{3} }{\chi } & -\frac{2 n
     p_{n} p_{\epsilon} \tau }{\chi } & \frac{2 p_{n} \left(c^2-n p_{n} \tau \right)}{\tau  \chi } \\
   -\frac{v_{3} \omega }{2} & \phi_{3} v_{1}  & \phi_{3} v_{2}   &
     c^2+\phi_{3} v_{3}  & -\phi_{3}  & -p_{n} v_{3} \\
   \frac{1}{4} (\omega -2 c  v_{1}) & \frac{1}{2} (c-\phi_{1} ) & -\frac{1}{2} \phi_{2}
       & -\frac{1}{2} \phi_{3}  & \frac{p_{\epsilon} \tau }{2} & \frac{p_{n}}{2}
     \\
  \end{array}
  \right]
\end{align*}
where $\phi_{i} = p_{\epsilon}\tau v_{i}$.

@@

>>>>

\documentclass[onecolumn]{aastex62}
\usepackage{times}
\usepackage{epsfig}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{color}
\usepackage{microtype}
\usepackage{float}
\usepackage{stfloats}
\usepackage{natbib}
\usepackage{verbatim}
\usepackage{placeins}
\graphicspath{{./figures/}}
\usepackage{wrapfig}
\hypersetup{backref,breaklinks,colorlinks,citecolor=blue}
\usepackage[all]{hypcap}
\usepackage{definitions}
\usepackage{subfiles}

@@

%\begin{abstract}
%\end{abstract}
\begin{abstract}
A problem of high importance in computational astrophysics is obtaining accurate
solutions to the Euler equations of hydrodynamics. We are interested in solving
the Euler equations in the context of core collapse supernovae (CCSNe).
The toolkit for high-order neutrino-radiation hydrodynamics (\thornado)
is being developed for CCSNe simulations and related problems utilizing a spatial discretization
based on the discontinuous Galerkin (DG) method.
Because the Euler equations form a hyperbolic set of partial differential equations,
they may be linearized to advection equations by the introduction of
chracteristic variables related to the eigen-decomposition.
This use of characteristic variables also increases the efficiency of the slope
limiting process. However, the addition of the nuclear equation of state equation
to the Euler equations makes the decompisition nontrivial.
We introduce the framework for the characteristic decomposition of the Euler equations
with the inclusion of the nuclear EOS terms and
present results from some initial tests.
\end{abstract}

@@

\section{Euler Equations of Gas Dynamics in Cartesian Coordinates}
\todo{Use this form of the Euler Equations?}
The non-relativistic Euler equations of gas dynamics
\citep[see, e.g.,][for details]{leveque:2002} in cartesian coordinates are given by
the equations of conservation of mass,
\beq
  \pderiv{\rho}{t} + \pderiv{}{x}(\rho u) = 0
  \label{eq:massConservation}
\eeq
conservation of momentum,
\beq
	\pderiv{(\rho u)}{t} + \pderiv{}{x}(\rho u^2 + p) = 0
  \label{eq:momentumConservation}
\eeq
conservation of energy,
\beq
  \pderiv{E}{t} + \pderiv{}{x}((E+p)u) = 0
  \label{eq:energyConservation}
\eeq
and conservation of \textcolor{red}{electron density}
\beq
  \pderiv{D}{t} + \pderiv{}{x}(D u) = 0
  \label{eq:electronConservation}
\eeq
where $\rho$ represents mass density, $u$ the fluid velocity, $p$ the fluid
pressure, $D=\rho y_e$ is the \textcolor{red}{what is this?},
and $E=\epsilon \rho +\frac{1}{2}\rho u^2$ the total specific energy (internal plus kinetic).
The system of equations is closed by the EOS which links the pressure to
the quantities evolved by Equations
\eqref{eq:massConservation}-\eqref{eq:electronConservation}. The inclusion of
Equation \eqref{eq:electronConservation} is because we require a nuclear EOS.
We may rewrite Equations \eqref{eq:massConservation}-\eqref{eq:electronConservation}
in a more convenient way:

@@

\subsection{Slope Limiting}
A common feature of numerical ODEs is unphysical oscillations in the solutions.
It is therfore of great interest in the DG algorithm to implement slope limiting
of the polynomial $\mathbf{U}_h$. We use the total variation diminishing (TVD)
slope limiter \citep[see, e.g.,][]{cockburn:1998} in conjunction with the
troubled cell indicator discussed in \citet{fu:2017} to prevent excessive limiting.
The general $s$-stage Runge-Kutta time stepping algorithm, including
the limiting process, can then be summarized as in \citet{cockburn:2001}:
\begin{itemize}
  \item[1.] Set $\bar{\vect{U}}^{(0)} = \bar{\vect{U}}^{n}$,
  \item[2.] For $i=1,\ldots,s$ compute:
  \begin{equation}
  \small  \bar{\vect{U}}^{(i)}
    = \Lambda^{\Pos}\Big\{\Lambda^{\TVD}\Big\{\sum_{j=0}^{i-1}\alpha_{ij}\,\bar{\vect{U}}^{(j)}+\beta_{ij}\,\Delta t\,\bar{\vect{F}}\big(\bar{\vect{U}}^{(j)}\big)\Big\}\Big\},
    \label{eq:rkStages}
  \end{equation}
  \item[3.] Set $\bar{\vect{U}}^{n+1}=\bar{\vect{U}}^{(s)}$.
\end{itemize}
Above, the TVD limiter and the `positivity' limiter preventing unphysical
states are denoted by the operators $\Lambda^{\TVD}\{\}$
and $ \Lambda^{\Pos}\{\}$, respectively. The SSP-RK3 coefficients $\alpha_{ij}$
and $\beta_{ij}$ may be found in Table~2.1 in \citet{cockburn:2001}.

@@

\beq
  \pderiv{\mathbf{F(\mathbf{U})}}{\mathbf{U}} =
  \mathcal{R} \Lambda \mathcal{R}^{-1},
\eeq
where the columns of $R$ contain the right eigenvectors of the Jacobian,
the rows of $\mathcal{R}^{-1}$ contain the left eigenvectors, and
$\Lambda$ is a diagonal matrix containing the eigenvalues of the Jacobian.
At this point, we will introduce the characteristic variable
$\mathbf{w} = \mathcal{R}^{-1}\mathbf{U}$. Multiplying both sides of Equation
\eqref{eq:charEq} by $\mathcal{R}^{-1}$, we are able to linearize the system of equations to
a system of advection equations

@@

%\begin{table*}[ht!]

\begin{align}
  \centering
	\pderiv{\mathbf{F}(\mathbf{U})}{\mathbf{U}}
	= \left[
		\begin{array}{cccccc}
			0 & 1 & 0 & 0 & 0 & 0 \\
			-v_{1}^{2} -p_{\tau}\tau^{2} - p_{\epsilon}\tau(\epsilon - \frac{v^{2}}{2}) & v_{1}(2-p_{\epsilon}\tau)  & -p_{\epsilon}v_{2}\tau & -p_{\epsilon}v_{3}\tau  & p_{\epsilon}\tau  & p_{n} \\
			-v_{1}v_{2} & v_2 & v_1 & 0 & 0 & 0 \\
			-v_{1}v_{3} & v_{3} & 0 & v_{1} & 0 & 0 \\
			v_{1}(-H - p_{\tau}\tau^{2} -p_{\epsilon}\tau(\epsilon - \frac{v^{2}}{2})) & H - p_{\epsilon}v_{1}^{2}\tau  & -p_{\epsilon}v_{1}v_{2}\tau & -p_{\epsilon}v_{1}v_{3}\tau  & v_{1}(1+p_{\epsilon}\tau) & v_{1} p_{n} \\
			-v_{1} Y_{e}/m_{B} & Y_{e}/m_{B} & 0 & 0 & 0 & v_{1} \\
		\end{array}
    \right]
\end{align}
%\caption{}
%\end{table*}

@@

\begin{align}
\Lambda =
\begin{bmatrix}
  v_{1} - c_{s} & 0 & 0& 0& 0& 0\\
  0 & v_{1} & 0 & 0 & 0 & 0 \\
  0 & 0 & v_{1} & 0 & 0 & 0 \\
  0 & 0 & 0 & v_{1} & 0 & 0 \\
  0 & 0 & 0 & 0 & v_{1} & 0 \\
  0 & 0 & 0 & 0 & 0 & v_{1} + c_{s}
\end{bmatrix}
\end{align}
where $c_{s} = \sqrt{\tau (n p_{n} + p p_{\epsilon}\tau - p_{\tau}\tau)}$ is
the local sound speed. The right eigenvectors are then given by
\begin{align*}
  \mathcal{R}_{1} =
  \left[
  \begin{array}{cccccc}
   1 & 0 & 1 & 1 & 0 & 1 \\
   v_{1}-c & 0 & v_{1} & v_{1} & 0 & c+v_{1} \\
   v_{2} & 1 & 0 & 0 & 0 & v_{2} \\
   v_{3} & 0 & 0 & 0 & 1 & v_{3} \\
   h-c v_{1} & v_{2} & \beta & 0 & v_{3} & h+c v_{1} \\
   n \tau  & 0 & 0 & \frac{\tau  \chi }{2 p_{n}} & 0 & n \tau  \\
  \end{array}
  \right]
\end{align*}
where the follow definitions have been used:
$h_{n} = \frac{c^2}{p_{\epsilon}\tau} + k $, $k = \frac{-n p_{n}
+ p_{\epsilon} (\frac{1}{2}v^2 + \epsilon) + p_{\tau}\tau}{p_{\epsilon}}$,
$\delta_{1} = v_{1}^{2}-v_{2}^{2}-v_{3}^{2}$,
$\chi = p_{\epsilon} ( \delta_{1} + 2\epsilon) + 2p_{\tau}\tau$, and
$\beta = \frac{1}{2} (\delta_{1}+2 \epsilon +\frac{2 p_{\tau} \tau }{p_{\epsilon}})$.
\begin{align*}
  \mathcal{R}_{1}^{-1} = \frac{1}{c^2}
  \left[
  \begin{array}{cccccc}
   \frac{1}{4} (2 c  v_{1}+\omega ) & \frac{1}{2} (-c- \phi_{1} ) & -\frac{1}{2} \phi_{2}
    & -\frac{1}{2} \phi_{3}  & \frac{p_{\epsilon} \tau }{2} & \frac{p_{n}}{2}
     \\
   -\frac{v_{2} \omega }{2} & \phi_{1} v_{2}  & c^2+\phi_{2} v_{2}  &
     \phi_{2} v_{3}  & -\phi_{2}  & -p_{n} v_{2} \\
   \frac{2 \chi  c^2+\alpha  \omega }{2 \chi } & -\frac{\phi_{1} \alpha  }{\chi } &
     -\frac{\phi_{2} \alpha  }{\chi } & -\frac{\phi_{3} \alpha }{\chi } &
     \frac{p_{\epsilon} \alpha  \tau }{\chi } & \frac{p_{n} \left(\alpha  \tau -2 c^2\right)}{\tau  \chi } \\
   -\frac{n p_{n} \omega }{\chi } & \frac{2 n p_{n} \phi_{1} }{\chi } & \frac{2 n p_{n}
     \phi_{2} }{\chi } & \frac{2 n p_{n} \phi_{3} }{\chi } & -\frac{2 n
     p_{n} p_{\epsilon} \tau }{\chi } & \frac{2 p_{n} \left(c^2-n p_{n} \tau \right)}{\tau  \chi } \\
   -\frac{v_{3} \omega }{2} & \phi_{3} v_{1}  & \phi_{3} v_{2}   &
     c^2+\phi_{3} v_{3}  & -\phi_{3}  & -p_{n} v_{3} \\
   \frac{1}{4} (\omega -2 c  v_{1}) & \frac{1}{2} (c-\phi_{1} ) & -\frac{1}{2} \phi_{2}
       & -\frac{1}{2} \phi_{3}  & \frac{p_{\epsilon} \tau }{2} & \frac{p_{n}}{2}
     \\
  \end{array}
  \right]
\end{align*}
where $\phi_{i} = p_{\epsilon}\tau v_{i}$.

@@

\appendix
